---
Title: Using custom functions for cleaner code
---

In the code above our `mutate` call was spanning more than twenty lines of code. This makes it
not only hard to see what's actually going on, it also makes debugging much more difficult.

In this case it is helpful to tuck the more complex parts of our code away in custom function.
Taking our last example, we could create two functions `update_min_form()`, that updates the model formula, and `lineaer_mod()`, that calculates a linear model and creates the data on th fly.

```{r}
# define custom functions
update_min_form <- function(upvars, dpvars) { # update_vars, dep_vars
    update(min_formula, # old formula
           reformulate(c(".", upvars), dpvars) # changes to formula 
           ) 
}

linear_mod <- function(form, prodvar, filter_inp) {
      lm(form,
  # create data on the fly
       data = filter(csat_named_top,
                     prodvar == "All" | .data$product == prodvar,
                     eval(filter_inp)
       )
    )
}
```

With both functions predefined, our original code becomes much clearer to read and easier to debug:

```{r}
all_grps_grid_final_res2 <- all_grps_grid_final |>
  rowwise() |>
  mutate(
    form = list2("{product}_{type}_{model_spec}_{dep_vars}" := 
                    update_min_form(update_vars, dep_vars)
  
    ),
    mod = list(linear_mod(form, product, filter_ls)),
    res = list(broom::tidy(mod)),
    modstat = list(broom::glance(mod))
  ) |>
  select(product:model_spec, dep_vars, mod:modstat)

all_grps_grid_final_res2
```
