---
output: html_document
title: "Mastering the Many Models Approach"
---

## Intro

* Tidyverse's "Many Models Approach" from R4DS simplifies running same model on different data subsets.

* This talk:
  (i) Update Many Models Approach with recent tidyverse syntax and
  (ii) introduce more helper functions for advanced use cases.

* Official Many Models Approach has been omitted from 2nd ed R4DS and is replaced by tidymodels.

* 'tidymodels' strong framework with definite advantages, but comes with conceptual and syntactic overhead.

* There is definitely room for an advanced Many Models Approach

Links:

R for Data Science (R4DS) first edition:
https://r4ds.had.co.nz

R for Data Science (R4DS) second edition:
https://r4ds.hadley.nz


## Setup

* We don't need all tidyverse packages: heavy lifting done by dplyr, tidyr and broom.

* dplyr version >= 1.0

```{r, warning=FALSE, message=FALSE}
library(dplyr)        # <- necessary
library(tidyr)        # <- necessary
library(broom)        # <- necessary
library(rlang)        # <- nice to have
library(modelsummary) # <- for saving output
library(purrr)        # <- not really needed
library(dplyover)     # <- only for the data
```

* Customer Satsifaction mock-up data from dplyover package

* Create clean data taking raw values from `csatraw` and recoded column names from `csat`:

```{r}
# create a look-up vector of old and new names
lookup_vec <- set_names(names(csatraw), names(csat))

# rename the columns
csat_named <- csatraw |>
  rename(any_of(lookup_vec)) |>
  select(cust_id, type, product, csat,
         ends_with("rating"))

glimpse(csat_named)
```

* Each row in the cleaned dataset csat_named represents a customer's response, with data on customer type, product type, and ratings on various touchpoints.


## Fundamentals

* The basic approach introduced in R4DS involves breaking down data into grouped subsets stored in a "nested" `data.frame`.

* The model fitting function is applied iteratively to each row and the results are appended to a new column.

* Results are processed and tidied, turning the model output into a list of small `data.frame`s and concatenating them into one large `data.frame`.

There are four essential components in the workflow above that we'll be discussing in detail below:

1.  nested data
2.  rowwise operations
3.  tidy results
4.  unnesting results


#### Nested data

* The Many Models Approach streamlines running models on data subsets, using a nested `data.frame` instead of a traditional base R for loop.

* `dplyr::nest_by(product)` creates a `data.frame` with one row for each product type

```{r}
csat_prod_nested <- csat_named |>
  nest_by(product) 

csat_prod_nested
```

* each row contains a `data.frame` for customers of the corresponding product type.

```{r}
csat_prod_nested$data[[1]]
```

#### Rowwise operations

* `nest_by()` groups data `rowwise()`, allowing dplyr operations to be applied on each row, useful when vectorized functions aren't available.

* A formula object, `base_formula`, defines the relationship between dependent and independent variables.


```{r}
base_formula <- csat ~ postal_rating + phone_rating + email_rating +
  website_rating + shop_rating

base_formula
```

* `mutate()` creates new columns, here, `mod` which contains our model

* `lm()` function is executed per row, creating a list-column, mod, with an `lm` object in each row.

```{r}
csat_prod_nested |>
  mutate(mod = list(lm(base_formula, data = data)))
```

* Caution when using `rowwise()`: wrapping in `list()` is necessary when the result is not an atomic vector and `ungroup()` is needed to revoke the row-by-row operation and avoid performance degradation.

#### Tidy results with broom

* Two functions from the 'broom' package, `broom::glance()` and `broom::tidy()`, are used to extract model statistics and regression terms respectively, yielding a nested data.frame.

```{r}
csat_prod_nested_res <- csat_prod_nested |>
  mutate(mod     = list(lm(base_formula, data = data)),
         modstat = list(broom::glance(mod)),
         res =     list(broom::tidy(mod)))

csat_prod_nested_res
```

#### Unnesting results

* `tidyr::unnest()` is used to convert list of `data.frame`s back into a regular `data.frame` for easy access to results.

* Model statistics for product subgroups, such as r-squared, p-value, and number of observations, are obtained by selecting product and `modstat` columns and unnesting the latter.

```{r}
csat_prod_nested_res |>
  select(product, modstat) |>
  unnest(modstat) |>
  select(r.squared, p.value, nobs)
```

* Coefficients, their size, and p-values are explored by selecting `product` and `res` columns and unnesting `res`, filtering out the intercepts.

```{r}
csat_prod_nested_res |>
  select(product, res) |>
  unnest(res) |>
  filter(term != "(Intercept)")
```

* The resulting data can be further manipulated, such as filtering non-significant coefficients or plotting results.

* Compare with the original 2017 syntax:

```{r}
csat_named |>
  group_by(product) |>
  nest() |>
  ungroup() |>
  mutate(mod     = map(data, ~ lm(base_formula, data = .x)),
         res     = map(mod, broom::tidy),
         modstat = map(mod, broom::glance))
```

* `map()` is used instead of `rowwise()` and `list()`, code clutters, when many variables are used with `pmap()`


## Extensions

* six advanced building blocks that help to tackle more complex use cases:

1.  create an overall category with `bind_rows()`
2.  add subgroups through filters with `expand_grid()`
3.  dynamically name list elements with `rlang::list2()`
4.  use data-less grids
5.  build formulas programmatically with `reformulate()`
6.  save model output to Excel with `modelsummary()`

#### Create an overall category with 'bind_rows()'

* To run an analysis on the entire dataset and different subsets simultaneously, `mutate()` and `bind_rows()` are used to create an additional overall product category.

```{r}
csat_all <- csat_named |>
  mutate(product = "All") |>
  bind_rows(csat_named) 

csat_all |> count(product)
```

* Now same workflow as above can be applied:

```{r}
csat_all |>
  nest_by(product) |>
  mutate(mod     = list(lm(base_formula, data = data)),
         res     = list(broom::tidy(mod)),
         modstat = list(broom::glance(mod)))
```


#### Add subgroups through filters with 'expand_grid()'

* Additional subgroups that meet specific filter criteria can be created using `expand_grid()`.

* To achieve this, we'll follow three steps:

1. A list of filter expressions, `filter_ls`, is created with elements that are either `TRUE` or unevaluated expressions for use later in `filter()`.

    ```{r}
    filter_ls <- list(
      All = TRUE,
      no_reactivate = expr(type != "reactivate")
    )

    filter_ls
    ```

* we use `expr()` to prevent evaluation

2. The nested `data.frame` is expanded for each filter category using `expand_grid()`, creating a tibble from all combinations of objects.

```{r}
csat_all_grps <- csat_all |>
  nest_by(product) |>
  expand_grid(filter_ls) |>
  mutate(type = names(filter_ls),
         .after = product)

csat_all_grps
```

* A new column, type, is added using `mutate()` to indicate the type of each row.

3.  We apply each filter to our data `rowwise` using `dplyr::filter(eval(filter_ls))`.

```{r}
csat_all_grps_grid <- csat_all_grps |>
  rowwise() |>
  mutate(data = list(
    filter(data, eval(filter_ls))
    ),
    .keep = "unused"
  )

csat_all_grps_grid
```

* Each filter in `filter_ls` is applied to the data rowwise using `dplyr::filter(eval(filter_ls))` within a `mutate()` call.

* From here the original workflow can be applied:

```{r}
csat_all_grps_grid <- csat_all_grps |>
  rowwise() |>
  mutate(mod = list(lm(base_formula, data = data)),
         res = list(broom::tidy(mod)),
         modstat = list(broom::glance(mod)))

csat_all_grps_grid |>
  select(product, type, modstat) |>
  unnest(modstat) |>
  select(-c(sigma, statistic, df:df.residual))

csat_all_grps_grid |>
  select(product, type, res) |>
  unnest(res) |>
  filter(term == "website_rating")
```

* Simple example, but can be scaled well.

#### Dynamically name list elements with 'rlang::list2()'

So far, we've wrapped the results of our `rowwise` operations in `list()` when they produced non-atomic vectors.

* A common issue when inspecting the results: elements are unnamed.

```{r}
csat_all_grps_grid$modstat[4]
```

* Enter `rlang::list2()`: dynamic unquoting of names helps name list-column elements for better tracking

* elements are named dynamically according to values in the product and type columns.

```{r}
csat_all_grps_grid <- csat_all_grps |>
  rowwise() |>
  mutate(mod     = list2("{product}_{type}" := lm(base_formula, data = data)),
         res     = list2("{product}_{type}" := broom::tidy(mod)),
         modstat = list2("{product}_{type}" := broom::glance(mod)))

csat_all_grps_grid$modstat[4]
```


#### Data-less grids

* A "data-less grid" is a more memory-efficient approach

* it's a nested `data.frame` *without* the data column, with combinations of subgroups created manually.

* Create an initial grid `all_grps_grid` of combinations between product and type, with an additional column of filter expressions.


```{r}
product <- c(
  "All", unique(csat_named$product)
)

all_grps_grid <- expand_grid(product, filter_ls) |>
  mutate(type = names(filter_ls),
         .after = product)

all_grps_grid
```

* Data subsets are generated on the fly in the `lm()` call by filtering the initial data `csat_named` based on product types and customer types.

1. Advanced filter expressions are used to handle cases where both the original data and the grid contain a column with the same name.

  `.env$product == "All" | .env$product == product`
  
  * `.env` is used to point to objects in enclosing environments
  
  * If the product in the grid (`.env$product`) is "All", all rows are selected
  
  * If the product is not "All", the `product` column in `csat_named` should match the value in the product column of the data-less grid.

2. Different customer types are filtered using the expressions stored in `filter_ls`:

```{r}
all_grps_grid_mod <- all_grps_grid |>
  rowwise() |>
  mutate(mod = list(
    lm(base_formula,
       data = filter(csat_named,
                     # 1. filter product categories
                     .env$product == "All" | .env$product == product,
                     
                     # 2. filter customer types
                     eval(filter_ls) 
                     )
       )
    )
    ) |>
  select(! filter_ls)

all_grps_grid_mod
```

* From here on the original workflow can be applied:

```{r}
all_grps_grid_res <- all_grps_grid_mod |>
  mutate(res     = list(broom::tidy(mod)),
         modstat = list(broom::glance(mod))) 

all_grps_grid_res |>
  select(product, type, modstat) |>
  unnest(modstat) |>
  select(-c(sigma, statistic, df:df.residual))
```


#### Build formulas programmatically with 'reformulate'

* `reformulate()` is a lesser-known base R function that generates formula objects based on character vectors, useful in the Many Models Approach.

```{r}
form1 <- y ~ x1 + x2
form1 

form2 <- reformulate(termlabels = c("x1", "x2"),
                     response = "y")
form2

identical(form1, form2)
```

* Two use cases for `reformulate()`: (1) Fitting multiple models to data in each row of a nested data frame with different predictor variables, and (2) Defining a base model in advance and adding different variables to it in each iteration.

1. A separate model for each independent variable is created by expanding the data-less grid with a character vector of independent variable names `indep_vars`, and using `reformulate(indep_vars, "csat")` instead of a formula object.

```{r}
indep_vars <- c("postal_rating",
                "phone_rating",
                "email_rating",
                "website_rating",
                "shop_rating")

all_grps_grid_vars <- all_grps_grid |>
   expand_grid(indep_vars)

all_grps_grid_vars
```

```{r}
all_grps_grid_vars_mod <- all_grps_grid_vars |>

  rowwise() |>

  mutate(mod = list(
    lm(reformulate(indep_vars, "csat"), # <- this part is new
       data = filter(csat_named,
                     .env$product == "All" | .env$product == product,
                     eval(filter_ls)
       )
    )
  )
  ) %>%
  select(! filter_ls)

all_grps_grid_vars_mod
```

2. A more powerful use case for `reformulate()` is to `update()` and add additional variables to a baseline model

* Create minimal model

```{r}
min_formula <- csat ~ postal_rating + phone_rating + shop_rating
min_formula
```

* The additional variables include `NULL` for the baseline model, one or more variable names and an interaction effect:

```{r}
update_vars <- list(base = NULL,
                    email = "email_rating",
                    website = "website_rating",
                    both = c("email_rating", "website_rating"),
                    both_plus_interaction = "email_rating*website_rating")
```

* Expand original grid with variables to update:

```{r}
all_grid_upd_vars <- all_grps_grid |>
  expand_grid(update_vars) |>
  mutate(model_spec = names(update_vars),
         .after = type)

all_grid_upd_vars
```

* An updated formula column, `form`, is created before applying `lm()`

```{r}
all_grid_upd_vars_form <- all_grid_upd_vars |>

  rowwise() |>

  mutate(
    form = list(
      update(
        min_formula, # old formula
        reformulate(c(".", update_vars)) # changes to formula
        )
      ),

    mod = list2( "{product}_{type}_{model_spec}" :=
    lm(form,
       data = filter(csat_named,
                     .env$product == "All" | .env$product == product,
                     eval(filter_ls)
                     )
       )
    )
  )
```

* Checking results:

```{r}
head(all_grid_upd_vars_form$mod, 5)
```

#### Save model output to Excel with 'modelsummary()'

* The `modelsummary()` function from the 'modelsummary' package is used to create an Excel file combining model statistics and estimator results, beneficial for reporting to non-R-users.

* `modelsummary()` accepts list-columns of model objects as input and supports various output formats.

* Additional arguments are used to trim results for a compact table, including omission of specific model statistics and the coefficient size of the intercept.

```{r, eval=FALSE}
# this saves the results to a `data.frame` in `out` and ...
# at the same time creates an .xlsx file
out <- modelsummary(models = all_grid_upd_vars_form$mod,
                    output = "model_results.xlsx",
                    gof_omit = "AIC|BIC|Log.Lik|RMSE",
                    coef_omit = "(Intercept)",
                    stars = TRUE,
                    statistic = NULL)
```

* The output includes model names indicating the subgroups of each `lm()` call.

* Position of p-value stars is suboptimal, workaround output as `data.frame`, `pivot_wider()` and then save to Excel:

```{r}
out <- modelsummary(models = all_grid_upd_vars_form$mod,
                    output = "data.frame",
                    gof_omit = "AIC|BIC|Log.Lik|RMSE",
                    coef_omit = "(Intercept)",
                    statistic = "stars") |>
  mutate(statistic = ifelse(statistic == "", "estimate", statistic)) |>
  select(-part) |>
  pivot_wider(names_from = statistic,
              values_from = -c(term, statistic),
              values_fn = as.numeric)

openxlsx::write.xlsx(out, "model_results2.xlsx")
```


## Endgame

* We can even compare two versions of the dependent variable by creating a character vector, `dep_vars`, holding the names of the dependent variables.

```{r}
csat_named_top <- csat_named |>
  mutate(csat_top = ifelse(csat == 5, 1, 0))

dep_vars <- c("csat", "csat_top")

all_grps_grid_final <- all_grid_upd_vars |>
  expand_grid(dep_vars)

all_grps_grid_final
```

* Next steps: the formula is updated, data is generated, the model is calculated, and results are prepared using 'broom'.

```{r}
all_grps_grid_final_res <- all_grps_grid_final |>

  rowwise() |>

  mutate(
    
  # dynamically name list
  form = list2( "{product}_{type}_{model_spec}_{dep_vars}" :=
  # update formula
    update(min_formula, # old formula
           reformulate(c(".", update_vars), dep_vars)) # changes to formula
  ),
    
  mod = list(
    lm(form,
  # create data on the fly
       data = filter(csat_named_top,
                     .env$product == "All" | .env$product == product,
                     eval(filter_ls)
       )
    )
  ),

  res = list(broom::tidy(mod)),

  modstat = list(broom::glance(mod))

  ) |>
  select(product:model_spec, dep_vars, mod:modstat)

all_grps_grid_final_res
```

* To keep code clean and facilitate debugging, complex parts can be tucked away in custom functions.

```{r}
# define custom functions
update_min_form <- function(upvars, dpvars) { # update_vars, dep_vars
    update(min_formula, # old formula
           reformulate(c(".", upvars), dpvars) # changes to formula 
           ) 
}

linear_mod <- function(form, prodvar, filter_inp) {
      lm(form,
  # create data on the fly
       data = filter(csat_named_top,
                     prodvar == "All" | .data$product == prodvar,
                     eval(filter_inp)
       )
    )
}

all_grps_grid_final_res2 <- all_grps_grid_final |>
  rowwise() |>
  mutate(
    form = list2("{product}_{type}_{model_spec}_{dep_vars}" := 
                    update_min_form(update_vars, dep_vars)
  
    ),
    mod = list(linear_mod(form, product, filter_ls)),
    res = list(broom::tidy(mod)),
    modstat = list(broom::glance(mod))
  ) |>
  select(product:model_spec, dep_vars, mod:modstat)

all_grps_grid_final_res2
```

* One possible way to refine results: filter models significant at the 10% level and arrange by adjusted r-squared in descending order:

```{r}
all_grps_grid_final_res |>
  unnest(modstat) |>
  select(-c(sigma, statistic, df:df.residual)) |>
  filter(p.value < 0.1) |>
  arrange(desc(adj.r.squared))
```

## FIN 

Link to full blog post::
https://tim-tiefenbach.de/post/2023-dplyr-many-models/

