---
Title: Ordering rows containing `NA`s
---
Base R's `order()` has an argument `na.last` which is set to `TRUE` by default and sorts `NA` to the bottom:
```{r}
mycars[order(mycars$mpg),]
```

Setting `na.last = FALSE` sorts `NA`s to the top: 
```{r}
mycars[order(mycars$mpg, na.last = FALSE),]
```

A bit surprisingly `na.last` can also be set `NA` which will remove `NA`s from the vector or `data.frame` making it a combined filter and ordering operation:
```{r}
mycars[order(mycars$mpg, na.last = NA),]
```

For some users this behavior might not be very intuitive, which is why splitting ordering and filtering operations is preferable for code clarity.

This is especially relevant for users coming from 'dplyr' or 'data.table'. `dplyr::arrange()` always orders `NA`s last and doesn't come with an option to change this behavior. `data.table::setorder()` has an argument `na.last`, but (1) it can only be set `TRUE` or `FALSE` and (2) it defaults to the former which is the opposite of what `base::order()` does:

```{r}
mycarsDT <- data.table::as.data.table(mycars)
data.table::setorder(mycarsDT, mpg)
mycarsDT
```
