---
Title: 'Expand: Passing a named list or data.frame to do.call'
---
The documentation of `order()` explicitly contains a warning, reminding us to `unname()` `data.frame`s 
that we pass to `do.call("order", df)`.

In most cases, forgetting to `unname()` a `data.frame` won't get us in trouble. If, however, one of
the column names corresponds to an argument of `order()`, that is `na.last`, `decreasing` or `method`, 
forgetting to `unname()` the `data.frame` will in most cases do something we don't expect, and 
probably even notice.

Let's create a small toy `data.frame` with three columns, `x`, `y` and `decreasing`, the latter indicating whether `x - y` is greater or less than zero:
```{r}
dat <- data.frame(x = c(3, 3, 6),
                  y = c(5, 2, 4),
                  decreasing = c(FALSE, TRUE, TRUE))
dat
```

When we want to order by all columns in ascending order, using the `do.call()` approach showed above,
we get the following result:
```{r}
dat[do.call("order", dat), ]
```

When `do.call()` is constructing the call, it is actually evaluating this call:
```{r, echo = FALSE}
call("order", dat)
```

As we can see, the third column is passed to `order()`s `decreasing` argument. It is interesting to
note that this doesn't seem to throw an error, although we order by two vectors and pass three Boolean 
values to `decreasing`.

When we `unname()` our `data.frame` before passing it to `do.call()`, we get the correct (and
slightly different) result:
```{r}
dat[do.call("order", unname(dat)), ]
```

Under the hood, this evaluates to the correct call:
```{r, echo = FALSE}
call("order", unname(dat))
```
