---
Title: How to safely use external inputs 
---

If we want to make sure that our `recode_col()` function uses the correct dictionary then we can
supply it as second argument:

```{r}
# recode function to be used within `dplyr::mutate(across(...))`
recode_col_safe <- function(x, dict) {
  
  recode_vec <- dict |>
    filter(col_nm == cur_column()) |>
    pull(label, name = value)
  
  dplyr::recode(x, !!! recode_vec)
}
```

In that case, it is not enough to only supply the bare function `recode_col` to `across`.
We need to create an anonymous function `\(x, dic) ... ` that calls `recode_col`. Here we can use
`.env$dat_dict` to tell dplyr to look for an object `dat_dict` in a parent environment (and not
inside our data.frame). This would prevent an error in case our data would contain an column with
the same name as our data dictionary `dat_dict`.

```{r}
# vector of columns to recode
cols_vec <- unique(dat_dict$col_nm)

# recoding defined columns
dat |> 
  mutate(across(all_of(cols_vec),
                \(x, dic) recode_col_safe(x, .env$dat_dict))
  )

```
