---
output:
  hugodownplus::md_document:
    use_boxes: TRUE
    toc: TRUE
# Documentation: https://sourcethemes.com/academic/docs/managing-content/

title: "Using a Data Dictionary to Recode Columns with dplyr"
subtitle: ""
summary: "This blog post provides an in-depth tutorial on using dplyr to recode and rename multiple columns according to a data dictionary."
authors: []
tags: ["R", "dplyr", "recode"]
categories: ["R", "dplyr", "recode"]
date: 2023-06-30
lastmod: 2023-06-30
featured: false
draft: false

# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image:
  caption: ""
  focal_point: ""
  preview_only: false

# Projects (optional).
#   Associate this post with one or more of your projects.
#   Simply enter your project's folder or file name without extension.
#   E.g. `projects = ["internal-project"]` references `content/project/deep-learning/index.md`.
#   Otherwise, set `projects = []`.
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

eframe <- function (..., .noWS = NULL, .renderHook = NULL) {
  htmltools:::validateNoWS(.noWS)
  contents <- rlang::dots_list(...)
  htmltools::tag("e-frame", contents, .noWS = .noWS, .renderHook = .renderHook)
}
```

## Intro

Today's blog post is all about recoding columns using a data dictionary and `dplyr::recode()`. 

Many datasets, especially from surveys, come along with a proper documentation often in form of
a so called "data dictionary". A data dictionary contains at least three pieces of information: 
the (i) column names that are used in the dataset as well as corresponding (ii) numeric values
and (iii) labels which translate those abstract numbers into meaningful terms.

At times, you may need to transform the raw values into their associated labels for tasks like reporting
or plotting. Here, `dplyr::recode()` serves as an efficient tool to programmatically recode, and also
rename, columns according to a data dictionary. 

Recently I revamped an old ETL script I had written, which recoded multiple datasets based on a
data dictionary. This script was from the pre-dplyr 1.0 era, so updating it provided a great opportunity
to revisit this task, this time armed with the latest dplyr version. The resulting approach was such a
significant improvement over my original method, that I felt compelled to share it here, serving as both
a personal reference and a resource for anyone confronted with similar challenges.

We'll start with a straightforward example to demonstrate the basic workflow. We then look at a more
advanced scenario involving multiple datasets and a comprehensive data dictionary. Lastly, this blog 
concludes with some thoughts about the recent changes in dplyr version 1.1.0 and `recode`'s new successor `case_match()`.


## Recoding one dataset

#### Setup

Lets start with a really simple dataset composed of three columms and fives rows:

```{r, warning=FALSE, message=FALSE}
library(dplyr)

# dataset
dat <- tibble(a = 1:5,
              b = c(10:14),
              c = c(20:24))
dat
```

Assume we have a data dictionary that looks like this:
```{r}
# dictionary
dat_dict <- tibble(
  
  col_nm = c(
    rep("b", 5),
    rep("c", 5)
  ),
  
  value = c(
    10:14,
    20:24
  ),
  
  label = c(
    letters[1:5],
    letters[6:10]
  )
)

dat_dict
```

The dictionary has three columns: `col_nm` records our dataset's column names, `value` holds the values,
as seen in our raw data `dat`, and `label` stores the corresponding labels.

Note, that our dataset contains a column `a` that is not part of the data dictionary, just to show
that the dictionary doesn't need to hold value-label pairs for all columns.

Typically, attached data dictionaries come in form of csv or xlsx files, but after reading them into 
R, they should be in a similar form as our mock dictionary above.

#### Preparations

Next, we need to prepare two things: (i) a custom function to recode a **single** column according to
the dictionary and (ii) a vector of columns names we want to recode. We will then use both inside
`dplyr::mutate(across(...))`.

Taking a closer look at our custom recode function, it accepts the column to recode, `x`, as its only
argument. The function works in two steps. Initially, we create a named vector of matching
label-value pairs. Next, we splice this vector as arguments into `dplyr::recode(x, ...)` using rlang's 
triple bang operator `!!!`.

```{r}
# recode function to be used within `dplyr::mutate(across(...))`
recode_col <- function(x) {
  
  recode_vec <- dat_dict |>
    filter(col_nm == cur_column()) |>
    pull(label, name = value)
  
  dplyr::recode(x, !!! recode_vec)
}
```

To create a named vector of matching label-value pairs we start with our data dictionary `dat_dict`.
We filter the current column using `cur_column()`, which is possible since we're going use this function
inside `dplyr::across()` where `cur_column()` yields us the string name of the current column. Finally,
we use `pull(label, name = value)` to get the labels as character vector, along with their matching 
values as names.

In the final step of our custom function, we "splice" this named vector of label-value pairs into 
`dplyr::recode()`. Despite "splicing" sounding rather technical, it's essentially an early evaluation.
We evaluate the vector `recode_vec` prior to processing the `dplyr::recode()` call. 
Assume we want to recode column `b` in `dat` and `recode_vec` looks like this:

```{r, eval=FALSE}
recode_vec <- c("10" = "a", "11" = "b", "12" = "c", "13" = "d", "14" = "e")
```

Then ...
```{r, eval=FALSE}
dplyr::recode(x, !!! recode_vec)
```

... becomes:
```{r, eval=FALSE}
dplyr::recode(x, c("10" = "a", "11" = "b", "12" = "c", "13" = "d", "14" = "e"))
```

Keep in mind, we're presuming hat the data dictionary is available in the global environment, and that our dataset
`dat` doesn't contain a column with an identical name as our data dictionary `dat_dict`. We can bolster
the safety of our approach by supplying the data dictionary as a second argument to our `recode_col()` function. 
For interested readers this is shown in the info box below.

::: {.info-box title="Expand: External inputs to custom functions used in `across`"}
```{r, child="external_inputs.Rmd"}
```
:::

With our custom recode function ready, the next thing we need is a vector of column names
that we want to apply this function to. A straightforward way to do this is to get all unique column
names from our data dictionary.

```{r}
# vector of columns to recode
cols_vec <- unique(dat_dict$col_nm)
cols_vec
```


#### Recoding

Now we are all set, and the only thing left to do is to call `across()` on all of the column names
in our data dictionary `all_of(col_vec)` and let our custom recode function `recode_col()` do its job.

```{r}
# recoding defined columns
dat |> 
  mutate(across(all_of(cols_vec),
                recode_col)
  )
```

#### Renaming columns

Occasionally, data dictionaries offer not just corresponding values and labels, but also new, typically
more descriptive, column names.

I've dedicated an entire <a href="../2022-rename-columns/" role="highlight" target="_blank">blog post</a>
to the subject of renaming columns based on a lookup table. However, as the approach is slightly different
when using a dictionary compared to a pure lookup table, and since this topic is quite relevant, let's
take a brief look at it.

Suppose our data dictionary, `dat_dict2` includes the original abbreviated column names `short_nm` and corresponding
descriptive column names `long_nm`:

```{r, echo=FALSE}
# dictionary
dat_dict2 <- tibble(
  
  short_nm = c(
    rep("b", 5),
    rep("c", 5)
  ),
  
  long_nm = c(
    rep("new_b", 5),
    rep("new_c", 5)
  ),
  
  value = c(
    10:14,
    20:24
  ),
  
  label = c(
    letters[1:5],
    letters[6:10]
  )
)
dat_dict2
```

::: {.output-box title="Expand to show code"}
```{r, eval=FALSE}
# dictionary
dat_dict2 <- tibble(
  
  short_nm = c(
    rep("b", 5),
    rep("c", 5)
  ),
  
  long_nm = c(
    rep("new_b", 5),
    rep("new_c", 5)
  ),
  
  value = c(
    10:14,
    20:24
  ),
  
  label = c(
    letters[1:5],
    letters[6:10]
  )
)
```
:::


This time, we only require a named vector of corresponding old and new column names. To create that
we use our data dictionary and filter it for distinct entries in `short_nm` and `long_nm`. Then we
use again `pull()` together with its `name` argument, but note that the old and new values are in reverse 
positions compared to `recode()`.

```{r}
rename_vec <- dat_dict2 |>
  distinct(short_nm, long_nm) |>
  pull(short_nm, name = long_nm)

rename_vec
```

Once we have this named vector of corresponding short and long column names we use `all_of()` inside
`rename()`:

```{r}
dat |>
  rename(all_of(rename_vec))
```

As we've demonstrated above, it's remarkably straightforward to not just recode, but also rename columns
according to a data dictionary. If your work solely involves single datasets, you can skip the next section,
which will expand upon the previous approach, demonstrating how to recode a list of datasets. 

## Recoding many datasets

#### Setup and Reasoning

Initially, the necessity of having a list of datasets and one comprehensive data dictionary may not
be evident. A plausible scenario, for instance, could be a customer survey program composed of numerous
surveys featuring similar, yet not identical, survey items. These survey results are stored in a generic
table within a database, with columns simply labeled `item1`, `item2`, etc. The number of survey items may
vary among surveys. Each row contains a respondent ID to identify a respondent and a survey ID to indicate
the specific customer journey under which a respondent was surveyed. As the surveys are similar but not
identical, the same column (e.g., `item1`) may contain different values across different surveys.
Even if column values are consistent, they could correspond to different labels.

Again, our objective is to recode—and while we're at it, also rename—all columns listed in the data
dictionary across all surveys.

To keep this example as straightforward as possible, we'll use a minimal setup. In addition to `dat` from
before, let's construct another small toy dataset `dat2` and nest both within a `data.frame` consisting
of two columns: the `id` of each dataset and the actual `data`.

```{r}
# another dataset
dat2 <- tibble(a = 1:5,
               d = 10:14,
               e = 7:11)


# a list of datasets
dat_ls <- tibble(id = c("dat1", "dat2"),
                 data = tibble::lst(dat, dat2))

dat_ls
```

This is how the "data" list-column looks like:

```{r}
dat_ls$data
```

We assume once more that we have a data dictionary, `dat_dict3`, which contains old, short column
names `short_nm`, new long column names `long_nm`, as well as a mapping between `value`s and `label`s.
The only difference from the previous example is that we now have an additional `id` column, indicating
to which dataset the value-label mappings belong.

```{r, echo=FALSE}
# a data dictionary containing codes for several datasets
dat_dict3 <- tibble(
  
  id = c(
    rep("dat1", 10),
    rep("dat2", 10)
  ),
  
  short_nm = c(
    rep("b", 5),
    rep("c", 5),
    rep("d", 5),
    rep("e", 5)
  ),
  
  long_nm = c(
    rep("new_b", 5),
    rep("new_c", 5),
    rep("new_d", 5),
    rep("new_e", 5)
  ),
  
  value = 
    c(
    10:14,
    20:24,
    10:14,
    7:11
  ),
  
  label = c(
    letters[1:5],
    letters[6:10],
    letters[6:10],
    17:21
  )
)
dat_dict3
```

::: {.output-box title="Expand to show code"}
```{r, eval=FALSE}
# a data dictionary containing codes for several datasets
dat_dict3 <- tibble(
  
  id = c(
    rep("dat1", 10),
    rep("dat2", 10)
  ),
  
  short_nm = c(
    rep("b", 5),
    rep("c", 5),
    rep("d", 5),
    rep("e", 5)
  ),
  
  long_nm = c(
    rep("new_b", 5),
    rep("new_c", 5),
    rep("new_d", 5),
    rep("new_e", 5)
  ),
  
  value = 
    c(
    10:14,
    20:24,
    10:14,
    7:11
  ),
  
  label = c(
    letters[1:5],
    letters[6:10],
    letters[6:10],
    17:21
  )
)
```
:::

#### Preparations

Before the actual recoding can take place, we have to prepare two functions: 

1. a custom function, `recode_col2()`, that recodes **one** column according to **a specified** dictionary and

1. another custom function, `recode_df()`, which applies `recode_col2()` inside `across()` to 
all specified columns.

The first function is pretty similar to what we have seen earlier, with the only difference being the
use of two arguments, the column to recode, `x` and the dictionary according to which the column should
be recoded `dict`:
```{r}
# recode function
recode_col2 <- function(x, dict) {
  
  col_nm <- cur_column()
  
  recode_vec <- dict |>
    filter(short_nm == cur_column()) |>
    pull(label, name = value)
  
  dplyr::recode(x, !!! recode_vec)
}
```

The second function, `recode_df`,  basically wraps the actual recoding that we've used in the section
above into a function. It takes two arguments: the dataset we want to recode, `dat`, and the `id` of
the dataset as specified in the data dictionary (which should be the same as in our nested data.frame `dat_ls`).

```{r}
recode_df <- function(dat, dat_id) {
  
  # 1. get the current entries 
  cur_dat_dict <- dat_dict3 |>
    filter(id == dat_id) 
  
  # 2. vector of columns to recode
  cols_vec <- unique(cur_dat_dict[["short_nm"]])
  
  # 3. use across with both inputs and recode_col2
  dat |> 
    mutate(across(all_of(cols_vec),
                  \(x) recode_col2(x, cur_dat_dict))
    )
}
```

This function will be applied iteratively to each individual dataset in our list of data.frames `dat_ls`.
The function consists of three steps:

1. We subset our data dictionary with the supplied data ID to ensure only the value-label mappings
of the current dataset remain. We call this subset of our dictionary `cur_dat_dict`.

1. We then create a vector of column names, `cols_vec`, that we want to recode. These will consist of 
all unique column names in the current dictionary `cur_dat_dict`.

1. Finally, we use `dplyr::across()` on `all_of` the columns in `cols_vec` and supply the current column
`x` and the current dictionary `cur_dat_dict` to `recode_col2()`.

#### Recoding

The last step is to iteratively apply our new function `recode_df()` to our nested data.frame `dat_ls`.
This requires us first to call `rowwise()` on `dat_ls`, which applies all subsequent dplyr functions
row-by-row. We then overwrite our column holding the `data` with `list(recode_df(data, id))`. 
It's important to note that our custom function must be wrapped in `list()`, as it returns a non-atomic vector (a 
list of `data.frames`).

```{r}
dat_ls |> 
  rowwise() |> 
  mutate(data = list(recode_df(data, id))) |>
  pull(data) # <= for better printing
```

#### Rename many datasets based on a data dicitonary

Analogous to `recode_df()`, we can create a function that renames all columns of a dataset, below called 
`rename_df()`. The function works in two steps. Initially, we create vector of old and new name pairs
based on the distinct entries of our data dictionary that are relevant for this dataset `filter(id == dat_id)`.
Next, we use this named vector within `rename(all_of()`:

```{r}
rename_df <- function(data, dat_id) {
  
  # 1. create a vector of old and new name pairs ...
  # ... based on the current dictionary
  rename_vec <- dat_dict3 |>
    filter(id == dat_id ) |>
    distinct(short_nm, long_nm) |>
    pull(short_nm, name = long_nm)

  # 2. use this vector on the supplied data
  data |>
    rename(all_of(rename_vec))
}
```

Applying this function to our nested `data.frame` of datasets is basically the same approach as 
outlined above:

```{r}
dat_ls |> 
  rowwise() |> 
  mutate(data = list(rename_df(data, id))) |>
  pull(data) # <= for betters printing
```



## Final Thoughts

Readers who follow me on Twitter might know that, although being a dplyr fan-boy, my feelings towards
`dplyr::recode()` were less enthusiastic. My main issue was the unconventional order of arguments,
which diverges from what we know from `rename()` or `mutate()`.

`r eframe(src="https://twitter.com/timteafan/status/1254898416402018310?s=61&t=i8eoMCFkLgkJKWoZHxVvOQ")`

The order of arguments was in part a reason why `recode()` was flagged as "questioning" in dplyr version
1.0. Since dplyr version 1.1.0 `recode()` moved one stage further in its life cycle and is now "superseded"
by `dplyr::case_match()`.

I'm not sure of the full implications of this development. If I understand the <a href="https://lifecycle.r-lib.org/articles/stages.html" role="highlight" target="_blank">life cycle stages</a> correctly, then "superseded" means that `dplyr::recode()` is not going away any time soon and will
continue to be maintained, though it will not see new features.

However, if there's a chance that `dplyr::recode()` might become deprecated in future major
releases, we would need to think about a workaround, since `case_match()` doesn't
support the splicing of named vectors as arguments that we have used above. In that case, I will 
certainly update this blog post.

Despite my initial skepticism towards `dplyr::recode()`, I have to concede that, particularly in
combination with `across()`, it provides a clear and straight-forward workflow. With its ability to be
used programmatically and to handel complex cases, I hope that this blog post has convincingly shown
the benefits of this approach.

My original dplyr workflow was much more convoluted. The curious reader can find it in the answers
to my <a href="https://stackoverflow.com/questions/56636417/bunch-recoding-of-variables-in-the-tidyverse-functional-meta-programing" role="highlight" target="_blank">question on StackOverflow</a> from a couple of years
ago. 

But even outside of dplyr I haven't encountered a similarly seamless approach to recoding multiple
columns across several datasets. If you are up for a challenge, I'd love to see what base R or
data.table solutions you can come with to tackle this problem. Let me know in the comments or via 
Twitter or Mastodon if you have an alternative approach.

`r hugodownplus::child_session_info(pkgs = "attached")`
